Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Proof-read and edited first half of paper in preparation for submission

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining spatial equity and accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[Some School]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[Another University]{Christopher D. Higgins\corref{2}}
   \ead{cd.higgins@utoronto.ca} 
    \author[Some School]{Antonio Paez\corref{2}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Public bicycle share programs are implemented to promote cycling as a
  convenient and sustainable mode of transportation. These systems can
  also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. Recent studies assessing equity in
  bike sharing have found differences in ridership or membership based on
  income, education, and trip purpose. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. In 2018, the
  non-profit organization that operates the program launched an equity
  initiative to provide subsidized memberships and to expand service by
  adding twelve docking stations. Previous research found that Hamilton's
  public bicycle share program targeted well disadvantaged areas in the
  city compared to programs in other Canadian cities. Since the time or
  distance that members of the population need to reach a bicycle share
  station decreases the potential of accessing the program, the location
  of stations matters. Unlike other public amenities with greater ability
  to adapt to crowding, the utility of stations is limited by the number
  of bicycles that they can hold, which makes crowding effects critical;
  the system may not necessarily be improved if stations are easy to reach
  but offer only a small number of bicycles. In this research, we revisit
  the case of Hamilton and investigate equity differentials in
  accessibility to bicycle share stations. We compare accessibility in the
  program with and without the equity stations to assess the effect of the
  initiative. Previous research on cycling equity or accessibility has
  used a macro-level approach with entire neighbourhoods or census tracts
  as the unit of analysis, although there are recent exceptions. In
  contrast, we implement our analysis parting from microscopic zones to
  better reflect travel to a docking station, which often happens at the
  sub-neighborhood level, and conduct a sensitivity analysis at several
  walking thresholds. We then reaggregate the estimated accessibility to
  dissemination areas for further analysis using census data. Equity
  analysis of this type is possible thanks to a newly developed approach
  for accessibility using balanced floating catchment areas. The addition
  of equity stations achieved its goal to increase accessibility for lower
  income populations, although the increase was relatively modest (2.10
  bicycles/person with equity stations vs 1.97 bicycles/person without
  equity stations). Further research is needed to determine whether this
  encouraged more cycling.
  \end{abstract}

@@

\hypertarget{background}{%
\section{Background}\label{background}}

Public bicycle share programs (PBSP) have been implemented in over 800
cities worldwide and a great deal has been learned about their typical
users (Fishman, 2016). In many cities, males use bike share more than
females (Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M. Wang, et
al., 2020; Winters et al., 2019) as do younger age cohorts (Brey et al.,
2017; Buck et al., 2013; Fuller et al., 2011). However, one study found
that bike share users in Washington, DC were more likely to be female
(Buck et al., 2013), which suggests that the gender gap among cyclists
who use bike share is less disparate than the gap for personal bicycle
use (Fishman, 2016). There is some evidence that bike share users are
less likely to own a car (Buck et al., 2013; Reilly, Noyes, et al.,
2020). However, the relationship between income or education and bike
share use is less clear. Stations in disadvantaged communities in
Chicago generated most of the average annual trips (Qian and Jaller,
2020) and individuals from minority or lower socioeconomic status
neighbourhoods in Minneapolis-St.~Paul used the city's PBSP more (J.
Wang and Lindsey, 2019a). Being university educated was also a
significant correlate of bike share use in Montreal, Canada (Fuller et
al., 2011). On the other hand, members who reside in
minority-concentrated and lower socioeconomic status neighbourhoods use
the Minneapolis-St Paul bike share more frequently (J. Wang and Lindsey,
2019a). Financial savings have been found to motivate those on a low
income to use bike share (Fishman, 2016). Many studies have found that
proximity to a bike share station, either living or working near one, is
an important determinant of use or having a membership (Fishman, 2016;
Fuller et al., 2011). This makes sense given that individuals are more
likely to use services or programs that they can easily access. Finally,
the potential of bike share programs to increase cycling levels has also
been explored recently (Hosford et al., 2018, 2019), but more
longitudinal evidence is needed to determine their impact on encouraging
more cycling and offering more opportunities for physical activity. On
the whole, these findings highlight the need for bike share systems to
be more highly accessible for diverse populations in order to increase
use beyond the ``typical'' users, which has been the focus of recent
research (Auchincloss et al., 2020; Hull Grasso et al., 2020; MacArthur
et al., 2020), and to offer more people the option of using sustainable
and active transportation.

Several studies have recently explored the equity of PBSP in North
American cities by primarily examining who has access to bike share
(e.g., differences by demographics or socioeconomic status) and where
stations are located. Equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) (Chen et al., 2019). Both are of interest to
researchers and transport planners since they are often linked in that
advantage, or conversely disadvantage, has spatial patterns. Using a
negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP among disadvantaged communities and found
some disparities. While annual members in disadvantaged communities have
a significantly lower share of trips compared to other areas in the
city, they make longer trips. This suggests that they may be using PBSP
for work commuting, which points to the importance of ensuring equitable
access (Qian and Jaller, 2020). Similar results were found in
Philadelphia, where lower income areas generate fewer trips and efforts
to increase equity within the program have not been as successful as
intended (Caspi and Noland, 2019). In the case of Seattle, all
neighbourhoods had some level of access to dockless bikes but those with
higher incomes and more residents of higher education had more bikes
(Mooney et al., 2019). Babagoli et al.~(2019) found that neighbourhoods
in New York City with higher affluence had the greater proportion of
Citi Bike stations. Overall, these findings suggest that horizontal
equity can be achieved while vertical equity is harder to attain.
Exploring transport equity by investigating where bicycle share stations
are located, often using neighbourhoods or census tracts as the spatial
unit of analysis, can ignore or miss the benefits that may be derived
from adjacent zones. Meaning that, stations may be lacking in certain
neighbourhoods but there may be stations accessible within a reasonable
walking cutoff time. This is where accessibility becomes an important
consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved since its conceptualization and become linked
with recent interest in prioritizing local proximity and modes that are
suitable for local trips like walking and cycling (Levine, 2020). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequities in usage, as well as guide interventions
that increase access for groups that are under-serviced or address gaps
in transportation options. It also addresses some of the challenges of
other performance measures such as level of service within a
transportation network by measuring person-based indicators and
exploring differences in use between population groups (Páez et al.,
2012).

In the context of PBSPs, accessibility refers to the distance an
individual must travel to reach a bicycle share station (Kabra et al.,
2020; J. Wang and Lindsey, 2019b). Since the time or distance that
members of the population need to reach a bicycle share station
decreases the potential of accessing the program and ultimately use of
the program, the location and size (e.g., maximum number of bicycles
available) of stations matters. Indeed, distance to bicycle share
stations is associated with use (Fuller et al., 2011; J. Wang and
Lindsey, 2019b) and can be a barrier to using PBSPs (Fishman et al.,
2014). Kabra et al.~(2020) found that the majority of bike share usage
in Paris comes from areas within 300m of stations. Furthermore, unlike
other public amenities with greater ability to adapt to crowding, like
health care services, the utility of stations is limited by the number
of bicycles that they can hold, which makes crowding effects critical.
The program may not necessarily be improved if stations are easy to
reach but offer only a small number of bicycles. Likewise, more people
may not opt to use the program if the supply of bicycles available at
the nearest station is insufficient to meet demand. The location and
size of stations is important to increase the utility of this public
transportation option for more people, thus achieving vertical equity.
Accessibility analyses for PBSPs constitute a positive and
evaluation-based approach that also has the potential to inform equity
efforts. For instance, Wang and Lindsey (2019b) investigated whether new
or relocated bicycle share stations increased accessibility and use,
which offered important insights to improve the performance of the
program.

Several approaches have been commonly used for measuring accessibility:
cumulative opportunities, gravity, and utility-based (Handy and
Niemeier, 1997). Paez et al (2012) provide a recent overview of various
formulations and applications of accessibility in transportation
research. Briefly, the gravity-based approach involves weighting
opportunities, for example the quantity of bicycle share stations, as
measured by a function of time, using a negative exponential form (Handy
and Niemeier, 1997). It takes both demand and congestion into account
(Páez et al., 2012). Floating catchment area (FCA) methods are a type of
gravity measure that have been used in healthcare accessibility research
but that are applicable to transportation case studies. This approach
involves producing flexible catchment areas for populations, recognizing
that people may be willing to travel to access particular services (Páez
et al., 2012). Thus, this model does a good job of considering potential
crowding of services using a binary distance function (Páez et al.,
2012). A recent improvement to this approach was achieved through a
simple and intuitive adjustment to the allocation of supply and demand,
which addressed the effects of demand and service inflation from the
conventional approach (see Paez et al., 2019).

Researchers have also taken different approaches when it comes to
aggregation of data, either by using the individual or household as the
smallest unit of analysis or larger spatial zones. Previous research on
bike share equity has typically used a macro-level approach with
aggregate data from entire neighbourhoods or census tracts (Babagoli et
al., 2019; Mooney et al., 2019; Qian and Jaller, 2020; J. Wang and
Lindsey, 2019a), although there are recent exceptions (Chen et al.,
2019; Chen and Li, 2021). This is also true for studies examining
correlates of bike share demand (J. Wang and Lindsey, 2019b). Handy and
Niemeier (1997) note that using disaggregated data in accessibility
analyses provides a more accurate estimate for individuals, which is
useful for addressing vertical inequities in PBSP usage.

Using balanced floating catchment area methods, a novel approach that
has not been used yet in cycling research, we examine accessibility to a
public bicycle share program in Hamilton, Ontario. Using disaggregate
population-level data, we (1) conduct a sensitivity analysis by
measuring accessibility and level of service to bike share stations at
different thresholds of walking to reach a station: 3 minutes, 5
minutes, 10 minutes, and 15 minutes; (2) explore the contribution of
specific stations that were added to Hamilton's PBSP to reducing both
horizontal and vertical inequities; and (3) examine whether disparities
in accessibility exist according to median household income of
dissemination areas within the core service area.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a
case study. The city launched a public bicycle share program, Hamilton
Bike Share, in March 2015 with 115 stations and 750 bicycles (Hamilton,
2015). Before June 2020, the program was known as SoBi Hamilton.
Stations are spaced between 300 and 600 metres apart (Scott et al.,
2021). The core service area spans 35 km\^{}2 of the city and roughly
138,000 can reach a bike share hub within 30 minutes of walking {[}see
Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This represents
roughly one fifth of the population in Hamilton, according to the 2016
Canadian Census. The program was enthusiastically welcomed in the city -
within three weeks of launching, 10,000 trips had been made (Hamilton,
2015). In 2017, Hamilton Bike Share Inc., the non-profit organization
that operates the program, initiated an equity program, Everyone Rides
Initiative (ERI), to remove barriers that may prevent individuals from
accessing bike share in Hamilton. An additional 75 bicycles and 15
stations were added to the program, which expanded it to more
disadvantaged areas in the core service area {[}see Figure 2{]}. The
program also offers subsidized memberships to individuals who identify
as low income. A comparable program can be found in Philadelphia (see
Caspi and Noland, 2019). As of June 2020, the bike share program has 900
bikes and 130 stations {[}see Figure
\ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
memberships (Hamilton, 2015).

Hamilton Bike Share has conducted one membership survey to date in 2018
(Civicplan, 2017), and the findings from 420 members are broadly in line
with the trends that were discussed above (see Fishman, 2016 for a
recent review of the literature). The majority of respondents live
within the core service area and the gender split is typical: 57\% of
respondents are male and 41\% are female. The majority of respondents,
both male and female, are between 25 and 34 years of age, but the
percentage of male respondents is higher in the subsequent age groups.
Respondents use bike share for commuting (40\% of trips) or errands and
meetings (24\% of trips), and nearly 50\% of trips have an average
length of 11 to 20 minutes. As a result of having a bike share
membership, 49\% of respondents report that they use their private
vehicle less often or much less often but 48\% report that their private
vehicle use has remained about the same. This suggests that SoBi
Hamilton has been useful for certain kinds of trips but not all, meaning
that some trips continue to require a private vehicle.

Further research has provided insights about the behaviour and route
choice of bike share users in Hamilton. The routes most frequently
travelled are longer than the shortest distance route from origin to
destination (Lu et al., 2018; Scott et al., 2021). Ridership is
influenced by weather conditions, temporal factors such as university
terms, employment, and proximity to important destinations like the
local university, but is not influenced by population and most
transportation infrastructure variables (Scott and Ciuro, 2019). Bike
share users prefer to deter from the shortest distance route to travel
on streets with cycling facilities or lower volumes of traffic (Lu et
al., 2018) and avoid steep slopes and busy roads (Scott et al., 2021).
It is important to note that these studies used daily ridership or GPS
data from trips taken before the equity stations were implemented, so
these findings are informative with respect to the conventional
stations.

Our analysis builds upon a previous and recent study (Hosford and
Winters, 2018), which found that disadvantaged areas in Hamilton are
better served by the city's PBSP compared to other Canadian cities
{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs
where advantaged areas have greater access. Hosford and Winters (2018)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton is by default in more
disadvantaged areas compared to other Canadian cities, but there is also
a great deal of variation in income in the city center as well because
of the local university and increasing gentrification in the downtown
area. In their analysis, Hosford and Winters (2018) did not
differentiate between the conventional stations and those added to
increase equity, which we refer to as equity stations in this paper.
Therefore, this paper continues their work and makes an important
contribution to the growing body of literature that examines equity
issues in PBSPs.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining spatial equity in accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[Some School]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[Another University]{Christopher D. Higgins\corref{2}}
   \ead{cd.higgins@utoronto.ca} 
    \author[Some School]{Antonio Paez\corref{2}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Public bicycle share programs (PBSPs) are implemented to promote cycling
  as a convenient and sustainable mode of transportation. These systems
  can also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. This system was
  expanded in 2018 by an equity initiative that added twelve docking
  stations with the explicit objective of increasing spatial equity in
  access. Since the cost of reaching a bicycle share station decreases the
  potential of accessing the program, the location of stations matters,
  and previous research found that Hamilton's public bicycle share program
  targeted well disadvantaged areas in the city. In this research, we
  revisit the case of Hamilton and investigate differentials in
  accessibility to bicycle share stations using a balanced floating
  catchment area (BFCA) accessibility approach. We compare accessibility
  with and without the equity stations to assess the effect of the
  initiative. We implement our analysis parting from micro zones to better
  reflect walking to a docking station, and conduct a sensitivity analysis
  at several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. The addition of
  equity stations increased the serviced population at every threshold
  examined, and although accessibility increased for the whole population,
  this increase was relatively modest especially for population in the
  bottom 20\% of median household income.
  \end{abstract}

@@

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These systems can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to lower income or disadvantaged populations. As
these systems become increasingly common, their introduction has been
accompanied by a flurry of research that investigates them from the
perspective of spatial and transportation equity (e.g., Smith et al.,
2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
these programs are available to the general public and ought to be
accessible to any individuals who wish to use them, research on PBSPs
located in North American and European cities indicates that there are
inequities in spatial accessibility in need of redress.

In this paper we investigate the case of the PBSP in the City of
Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
as Social Bicycles, or SoBi) was launched in 2015 and currently has over
900 operational bicycles and 130 docking stations. An interesting
feature of this system is that it was expanded by an equity initiative
that introduced twelve docking stations in 2018 with the explicit
objective of increasing spatial accessibility. Hamilton Bike Share was
already studied by Hosford and Winters (2018) as part of a selection of
PBSPs in Canada. These authors found that most of the cities in their
research could benefit of greater efforts to expand service to areas
with lower socio-economic populations - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research is the use of a balanced floating catchment area (BFCA)
accessibility approach that accounts for the supply of stations and the
potential demand from the population serviced. This requires a more
disaggregated approach than the use of census geographies (q.v., Hosford
and Winters, 2018), since many trips to a bike share station likely
happen at a level lower than even the smallest census geography. For
this reason, we implement our analysis parting from micro population
zones to better reflect the friction of walking to a docking station.
Further, we conduct a sensitivity analysis at several walking time
thresholds. In terms of spatial equity, we compare accessibility with
and without the equity stations to assess the effect of the initiative,
before reaggregating the data for further analysis using median
household income information from the census. We find that the addition
of equity stations does increase the serviced population at every
threshold examined. However, although accessibility increased for the
whole population, this increase was relatively modest, especially for
population in the bottom 20\% of median household income. The analysis
also identifies areas under-serviced within the core service area of the
system.

This paper is an example of open and reproducible research that uses
only open software for transportation and statistical analysis (Bivand,
2020; Lovelace, 2021). All data were obtained from publicly available
sources and organized in the form a data package. Following best
practices in spatial data science (Brunsdon and Comber, 2020), the code
and data needed to reproduce (or modify/extend) the analysis are
available for download.\footnote{https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing}

\hypertarget{literature-review}{%
\section{Literature Review}\label{literature-review}}

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
({\textbf{???}}; Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M.
Wang, et al., 2020; Winters et al., 2019) as do younger age cohorts
(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
one study found that bike share users in Washington, DC were more likely
to be female (Buck et al., 2013), which suggests that the gender gap
among cyclists who use bike share is less disparate than the gap for
personal bicycle use (Fishman, 2016). There is some evidence that bike
share users are less likely to own a car (Buck et al., 2013; Reilly,
Noyes, et al., 2020). However, the relationship between income or
education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighbourhoods in
Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
2019a). Similar findings were reported in London ({\textbf{???}}). Being
university educated was a significant correlate of bike share use in
Montreal, Canada (Fuller et al., 2011). Perhaps not coincidentally,
financial savings have been found to motivate those on a low income to
use bike share (Fishman, 2016).

Many studies have found that geographic proximity to a bicycle share
station is an important determinant of membership and use (Fishman,
2016; Fuller et al., 2011). This makes sense given that individuals are
more likely to use services or programs that they can easily reach.
Several studies have recently explored the equity of PBSPs by primarily
examining who has access to bike share (e.g., differences by
demographics or socioeconomic status) and where stations are located. It
is important to note that equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) (Chen et al., 2019). Both are of interest to
researchers and transport planners since they are often linked in that
advantage, or conversely disadvantage, has spatial patterns. Using a
negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP among disadvantaged communities and found
some disparities. While annual members in disadvantaged communities have
a significantly lower share of trips compared to other areas in the
city, they make longer trips. This suggests that they may be using PBSPs
for utilitarian trips (e.g., commuting), which points to the importance
of ensuring equitable access (Qian and Jaller, 2020). Similar results
were found in Philadelphia, where lower income areas generated fewer
trips and it was suggested that efforts to increase equity within the
program have not been as successful as intended (Caspi and Noland,
2019). In the case of Seattle, all neighbourhoods had some level of
access to dockless bikes but those with higher incomes and more
residents of higher education had more bikes (Mooney et al., 2019).
Babagoli et al.~(2019) also found that neighbourhoods in New York City
with higher affluence had the greatest proportion of Citi Bike stations.
Researchers have recommended targeted expansion of stations to areas
that use bike share but have inequitable access ({\textbf{???}}).

On the whole, existing studies highlight the need for bicycle share
systems to be more highly accessible for diverse populations in order to
increase use beyond the ``typical'' users. This has been the focus of
recent research (see, among others, Auchincloss et al., 2020; Hull
Grasso et al., 2020; MacArthur et al., 2020). Offering more people the
option of using sustainable and active transportation, particularly
those who have lower socioeconomic status and might benefit the most, is
a worthy policy goal for cities with PBSPs. Several cities have launched
specific programs to address barriers to bike share or to expand service
to more deprived areas: London, United Kingdom ({\textbf{???}});
Philadelphia, Pennsylvania (Caspi and Noland, 2019), and Hamilton,
Ontario (Hosford and Winters, 2018). Findings to date suggest that
horizontal equity can be easier to achieve than vertical equity.
However, exploring transportation equity by investigating where bicycle
share stations are located, often using neighbourhoods or census tracts
as the geographical unit of analysis, can ignore or miss the benefits
that may be derived from adjacent zones. Meaning that, stations may be
lacking in certain neighbourhoods but there may be stations accessible
within a reasonable walking time. This is where geographical
accessibility becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent interest in prioritizing local proximity and modes
that are suitable for local trips like walking and cycling (Levine,
2020). As a measure of the ease of reaching potential destinations
spread spatially in a given area, accessibility is relevant to PBSPs
because it can identify current inequities in the provision of
infrastructure, as well as guide interventions that increase access for
groups that are under-serviced or address gaps in transportation
options. It also addresses some of the challenges of other performance
measures such as level of service within a transportation network by
measuring person-based indicators and exploring differences in use
between population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., maximum number of bicycle racks available) of
stations matters. Indeed, distance to bicycle share stations is
associated with use (Fuller et al., 2011; J. Wang and Lindsey, 2019b)
and can be a barrier to using PBSPs (Fishman et al., 2014). Kabra et
al.~(2020) found that the majority of bike share usage in Paris comes
from areas within 300 metres of stations, which amounts to 2-4 minutes
walking by an able-bodied adult. Furthermore, similar to other public
amenities affected by crowding, like health care services (e.g.,
Pereira, Carlos Kauê Vieira Braga, et al., 2021), the utility of
stations is limited by the number of bicycles that they can hold. The
program may not necessarily be improved if stations are easy to reach
but offer only a small number of bicycles. Likewise, more people may not
opt to use the program if the supply of bicycles available at the
nearest station is insufficient to meet demand. The location and size of
stations is important to increase the utility of this public
transportation option for more people, thus achieving vertical equity.
Accessibility analyses for PBSPs constitute a positive and
evaluation-based approach that also has the potential to inform equity
efforts. For instance, Wang and Lindsey (2019b) investigated whether new
or relocated bicycle share stations increased accessibility and use,
which offered important insights to improve the performance of the
program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). Geurs and Van Wee
(2004) and Paez et al.~(2012) provide recent overviews of various
formulations and applications of accessibility in transportation
research. The common gravity-based approach for example involves
weighting destination opportunities, such as the quantity of bicycle
share stations, by the time required to reach them from an origin using
an impedance function (Handy and Niemeier, 1997; Kwan, 1998). However,
while such measures are suitable for capturing the potential for
reaching destinations from a given location, they do not take demand or
congestion effects into account.

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centres and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, {\textbf{???}};
{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
recent improvement to this approach was achieved through a simple and
intuitive balancing of the impedance that addressed the effects of
demand and service inflation found in previous FCA approaches (see Paez
et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregate data from entire
neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequities in PBSP usage.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregate population-level data, in this paper we examine
accessibility to a public bicycle share program in Hamilton, Ontario. We
(1) conduct a sensitivity analysis by measuring accessibility and level
of service to bike share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median household income of dissemination areas within the core service
area.

@@

\hypertarget{original-system}{%
\subsection{Original System}\label{original-system}}

The focus of this study is the city of Hamilton, located in Ontario,
Canada. The city launched a public bicycle share program in March 2015
with 115 stations and 750 bicycles (Hamilton, 2015). Before June 2020,
the program was known as Social Bicycles or SoBi Hamilton but is now
called Hamilton Bike Share. Stations are spaced between 300 and 600
metres apart (Scott et al., 2021). The core service area spans 40 sq.km
of the city although it was planned to be 20 sq.km ({\textbf{???}}), and
roughly 138,000 people are within 30 minutes walking of a bike share
station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
represents roughly one fifth of the total population of the Hamilton
Census Metropolitan Area according to the 2016 Canadian Census. The City
of Hamilton undertook a large public engagement campaign to validate the
locations of bicycle share stations that had been selected and to
crowdsource potential locations for additional locations
({\textbf{???}}). Most of the suggested locations were in the east end
of the core service area that lack stations or in neighbourhoods not
serviced by the PBSP. The program was enthusiastically welcomed in the
city in 2015 - within three weeks of launching, 10,000 trips had been
made (Hamilton, 2015), however inadequate coverage given the size of the
Hamilton Bike Share's service area was identified as a problem early on,
and transportation planners noted that the small size (i.e., supply of
bicycle racks) and low quantity of stations would lead to challenges in
balancing the system ({\textbf{???}}). Based on public feedback, 13
stations were added to the original system.

\hypertarget{equity-initiative}{%
\subsection{Equity Initiative}\label{equity-initiative}}

In 2017, Hamilton Bike Share Inc., the non-profit organization that
operates the program, initiated an equity program, Everyone Rides
Initiative (ERI), with the objective of reducing barriers that may
prevent individuals from accessing bike share in Hamilton. Additional
bicycles and stations were added to the program which expanded it to
more disadvantaged areas in the core service area {[}see Figure 2{]}.
The program also offers subsidized memberships to individuals who
identify as low income, and complements these initiatives with cycle
skills education. A comparable program can be found in Philadelphia (see
Caspi and Noland, 2019).

\hypertarget{current-system}{%
\subsection{Current System}\label{current-system}}

As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations {[}see
Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
memberships (Hamilton, 2015). The core service area remains 40 sq.km.
The PBSP has 13 stations that were added as part of the ERI initiative;
we refer to these as ``equity stations'' throughout the paper while all
other stations are ``conventional stations''. In total, 117 stations are
``conventional'' and 13 are ``equity''.

\hypertarget{membership}{%
\subsection{Membership}\label{membership}}

Hamilton Bike Share conducted one membership survey in 2018 (Civicplan,
2017) and the findings from 420 members are broadly in line with the
trends that were discussed above (see Fishman, 2016 for a recent review
of the literature). The majority of respondents live within the core
service area and the gender split is expected: 57\% of respondents are
male and 41\% are female. The majority of respondents, both male and
female, are between 25 and 34 years of age, but the percentage of male
respondents is higher in the subsequent age groups. Respondents use bike
share for commuting (40\% of trips) or errands and meetings (24\% of
trips) and nearly 50\% of trips have an average length of 11 to 20
minutes. As a result of having a bike share membership, 49\% of
respondents report that they use their private vehicle less often or
much less often but 48\% report that their private vehicle use has
remained about the same. This suggests that Hamilton Bike Share has been
useful for certain kinds of trips but not all, meaning that some trips
continue to require a private vehicle.

\hypertarget{relevant-research}{%
\subsection{Relevant Research}\label{relevant-research}}
